---
format: revealjs
---

## Today's Agenda {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

```{r}
# background-size="1920px 1080px"
library(tidyverse)
library(readxl)

# Input data
# Add a corrected "year" because FH "Edition" refers to the reporting year and not the year being coded
d <- read_excel("../../Data_to_Explore/Freedom_House/Freedom_House_US_Data-1973-2024.xlsx") |>
  mutate(
    year = Edition - 1
  )
```

<br>

::: {.r-fit-text}

**I. Road testing measures of democracy**

:::

- Forecasting the future of American "freedom" using FH

- What could government do to improve American "freedom"?

<br>

::: r-stack
Justin Leinaweaver (Fall 2024)
:::

::: notes
Prep for Class

1. Review Canvas submissions

2. Readings: [Freedom House Methodology](https://freedomhouse.org/reports/freedom-world/freedom-world-research-methodology)

3. Data: Freedom_House_US_Data-1973-2024.xlsx

<br>

Let's warm up with election news!

- *Open google news and search 'election'*

- **What's been going on since we last met? Anything noteworthy?**

- **Any changes in the stakes we discussed last class?**

:::


## FH's "Freedom in the World" {background-image="Images/background-white_house_graphic_blue_v2.png" .smaller .center}

<br>

:::: {.columns}

::: {.column width="47%"}

<br>

<br>

A useful argument about the quality of any measure of democracy should consider the **definition**, data **sources**, the **instrument** and the measurement **process**.
:::

::: {.column width="6%"}

:::

::: {.column width="47%"}

**Freedom House Total Score**

- PR + CL

**Political Rights (PR)**

- A. Electoral Process (x3)
- B. Pluralism & Participation (x4)
- C. Functioning of Government (x3)

**Civil Liberties (CL)**

- D. Freedom of Expression & Belief (x4)
- E. Associational Rights (x3)
- F. Rule of Law (x4)
- G. Personal Autonomy & Rights (x4)

:::

::::

::: notes

Specifically this week we've been focused on the Freedom House "Freedom in the World" measures

<br>

**As a refresher, what strengths and weaknesses did we highlight when discussing the FH approach to measuring 'freedom'?**

- **e.g. definition, sources, instrument and process**

<br>

**Are we convinced or not that these 'freedom' measures can be used in place of measuring democracy around the world? Why or why not?**

:::



## Forecasts of American "Freedom" {background-image="Images/background-white_house_graphic_blue_v2.png" .center}

:::: {.columns}

::: {.column width="50%"}
```{r, fig.retina=3, fig.align='center', fig.asp=.9, fig.width=6}
d |>
  filter(!is.na(PR)) |>
  pivot_longer(cols = c(A, B, C), names_to = "Index", values_to = "Value") |>
  ggplot(aes(x = year, y = Value, color = Index)) +
  geom_hline(yintercept = c(12, 16), linetype = "dashed", color = "grey") +
  geom_line() +
  theme_bw() +
  #facet_wrap(~ Index, ncol = 4) +
  labs(x = "", y = "Index Scores",
       caption = "Source: Freedom House (2024)",
       title = "Political Rights Categories") +
  scale_x_continuous(breaks = seq(2012, 2028, 2), limits = c(2012, 2028)) +
  scale_y_continuous(breaks = seq(2, 16, 2), limits = c(1, 16)) +
  guides(color = FALSE) +
  annotate("text", x = 2012, y = c(9.5, 10.6, 15.5), label = c("C. Functioning of Government", "A. Electoral Process", "B. Pluralism & Participation"), size = 3, hjust = 0, color = c("#7570B3", "#1B9E77", "#D95F02")) +
  scale_color_brewer(palette = "Dark2") +
  annotate("polygon", x = c(2023, 2028, 2028), y = c(14, 10, 16), fill = "#D95F02", alpha = .2) +
  annotate("polygon", x = c(2023, 2028, 2028), y = c(10, 8, 12), fill = "#1B9E77", alpha = .2) +
  annotate("polygon", x = c(2023, 2028, 2028), y = c(9, 7, 11), fill = "#7570B3", alpha = .1)
```
:::

::: {.column width="50%"}
```{r, fig.retina=3, fig.align='center', fig.asp=.9, fig.width=6}
d |>
  filter(!is.na(PR)) |>
  pivot_longer(cols = c(D,E,F,G), names_to = "Index", values_to = "Value") |>
  ggplot(aes(x = year, y = Value, color = Index)) +
  geom_hline(yintercept = c(12, 16), linetype = "dashed", color = "grey") +
  geom_line() +
  theme_bw() +
  #facet_wrap(~ Index, ncol = 4) +
  labs(x = "", y = "Index Scores",
       caption = "Source: Freedom House (2024)",
       title = "Civil Liberties Categories") +
  scale_x_continuous(breaks = seq(2012, 2028, 2), limits = c(2012, 2028)) +
  scale_y_continuous(breaks = seq(2, 16, 2), limits = c(1, 16.5)) +
  scale_color_brewer(palette = "Dark2") +
  guides(color = FALSE) +
  annotate("text", x = 2012, y = c(10.5, 13.5, 14.5, 16.5), label = c("E. Associational Rights", "F. Rule of Law", "G. Personal Autonomy", "D. Freedom of Expression"), size = 3, hjust = 0, color = c("#D95F02", "#7570B3", "#E7298A", "#1B9E77")) +
  annotate("polygon", x = c(2023, 2028, 2028), y = c(14, 11, 16), fill = "#E7298A", alpha = .1) +
  annotate("polygon", x = c(2023, 2028, 2028), y = c(14, 11, 16), fill = "#1B9E77", alpha = .1) +
  annotate("polygon", x = c(2023, 2028, 2028), y = c(11, 8, 16), fill = "#1B9E77", alpha = .2) +
  annotate("polygon", x = c(2023, 2028, 2028), y = c(11, 8, 12), fill = "#D95F02", alpha = .1)
```
:::

::::

::: notes

Alright, let's use these Freedom House instruments to guide us in developing forecasts for the future.

<br>

Groups, touch base on your forecasts for a second Trump administration and get ready to report back

- *PRESENT and DISCUSS changes in specific indicators*

<br>

Groups, touch base on your forecasts for a first Harris administration and get ready to report back

- *PRESENT and DISCUSS changes in specific indicators*

<br>

**Bottom line, do these seven indicators represent a significant difference between the two presidential candidates? Why or why not?**

:::



## {background-image="Images/background-white_house_graphic_blue_v2.png" .center}

### What is a specific policy you would like to see the US government implement that would make a significant, positive impact on American freedom moving forward?

::: notes

*PRESENT and DISCUSS each*

:::




## For Next Class {background-image="Images/background-blue_triangles_flipped.png" .center}

<br>

::: {.r-fit-text}
**Evaluating the V-Dem egalitarian democracy index**

1. Compare this new measure to the prior three

2. Reflect on the role of equality in democracy

:::

::: notes

Next class we shift to our final research project that aims to measure democracy

- The V-Dem egalitarian democracy index will ask us to consider the role of equality as a necessary condition of democracy

<br>

The two prompts are on Canvas

- **Questions on the assignment?**

:::
