---
format: revealjs
---

## Today's Agenda {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

```{r}
# background-size="1920px 1080px"
library(tidyverse)
library(readxl)

# Input ANES data
d <- read_excel("../../Data_to_Explore/ANES/ANES-1978-2020-Issue_Polarization_Variables-PROCESSED.xlsx", na = "NA", guess_max = 100000) |>
  mutate(
    Age = case_when(
      VCF0103 == 1 ~ "18-29",
      VCF0103 == 2 ~ "30-45",
      VCF0103 == 3 ~ "46-61",
      VCF0103 == 4 ~ "62-77",
      VCF0103 == 5 ~ "78-93",
      VCF0103 == 6 ~ "94-109",
      VCF0103 == 7 ~ "110-125"
    )
  )

d2 <- d |> filter(VCF0004 == 2020)
```

<br>

::: {.r-fit-text}

**II. Contemporary Threats to American Democracy**

- Using ANES data to measure issue polarization

:::

<br>

::: r-stack
Justin Leinaweaver (Fall 2024)
:::

::: notes
Prep for Class

1. Readings
    - Aldrich 2017 on ANES: "The American National Elections Study as 'Gold Standard' for Survey Research in the Twenty-First Century"
    
    - Review the ANES data extract and codebook: Make sure you understand what each variable represents and how they are coded

<br>

Today I want us to get hands-on with some high quality data on the American electorate

<br>

Specifically, I've given you two extracts from the American National Election Studies surveys

- The "original" file is the variables as they appear in the raw data

- The "processed" version is me recoding the levels to help you understand what is being measured in each variable

<br>

SLIDE: Before analyzing, let's discuss the source of this data

:::



## {background-image="Images/background-white_house_graphic_blue_v2.png" .center}

::: {.r-fit-text}
**The American National Election Study (ANES)**
:::

<br>

::: {.r-fit-text}

**Big Picture Goals of the Project**

1. Analyzing individuals as voters

2. Analyzing the public's role in our democracy

:::

::: notes

The individual goals of the project are meant to help us answer two questions 

1. Why do some people vote and other do not?

2. How do voters decide who/what to vote for?

<br>

The second big picture goal is meant to help us understand the mass public more broadly

- If democracy depends on voters using their vote to punish bad leaders and reward good ones, does the data show us actually doing that or not?

<br>

**Generally speaking, did the write-up by Aldrich claiming the ANES is a "gold standard" give you confidence in the quality of the data?**

- **What specific design choices stood out to you as boosting your confidence?**

<br>

In our work last class, we examined polarization trends in the Fiorina and Abrams (2008)

- That research only had data up to 2007 so today we will extend those analyses to 2020 using the latest ANES data

- SLIDE

:::



## Party Identification in the ANES {background-image="Images/background-white_house_graphic_blue_v2.png" .center}

<br>

### Visualize the breakdown of respondents by party identification (VCF0303) across time

- e.g. the proportion of Democrats, Republicans and Independents across time

::: notes

Lots of options for how to do this

- RTutor can do it for you!

- In Excel you'd need a couple of steps:
    1. Make a pivot table that counts the responses per year, 
    2. Add a column to convert the counts to proportions (e.g. count/total)
    3. Highlight the table and insert a line plot

- Get to it!

<br>

SLIDE: The result

:::



## Party Identification in the ANES {background-image="Images/background-white_house_graphic_blue_v2.png" .center}

```{r, fig.retina=3, fig.align='center', fig.asp=.618}
d |>
  filter(!is.na(VCF0303)) |>
  group_by(VCF0004) |>
  summarize(
    Democrats = mean(VCF0303 == "Democrats"),
    Republicans = mean(VCF0303 == "Republicans"),
    Independents = mean(VCF0303 == "Independents")
  ) |>
  pivot_longer(cols = Democrats:Independents, names_to = "party", values_to = "value") |>
  ggplot(aes(x = VCF0004, y = value, color = party)) +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(0, .7), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1978, 2020, 2)) +
  labs(x = "", y = "",
       title = "The partisan makeup of the American electorate has been fairly stable since 1978") +
  scale_color_manual(values = c("blue", "darkgrey", "red")) +
  guides(color = "none") +
  annotate("text", x = 1980, y = c(.09, .27, .49), label = c("Independents", "Republicans", "Democrats"), color = c("darkgrey", "red", "blue"))
```

::: notes

I actually think the relative stability of these proportions is useful for our work today measuring polarization

- If party identification is stable but issue positions are moving then we have stronger evidence of polarization!

:::




## Polarization and the ANES {background-image="Images/background-white_house_graphic_blue_v2.png" .center}

<br>

::: {.r-fit-text}
1. Gay Adoption (VCF0878, p388)

2. Limit trade to protect US jobs (VCF9231, p600)

3. Death penalty for murderers (VCF9237, p603)

4. Guns (VCF9238, p604)
:::

::: notes

The ANES includes a bunch of survey questions designed to track issue positions across time

- HOWEVER, not all of them are used in each survey and some have more complicated response options than others

- I've chosen these four because they have been asked many times in the past, their response options are fairly simple and they represent big divides between the parties at least at the elite level

<br>

Note that the "limiting trade" one is particularly relevant today given that the most common approach to limiting imports is by imposing tariffs!

<br>

*Assign each to GROUPS (mix up the groups as you need a mix of skill levels with stats in each one!)*

<br>

GROUPS, review the question in the codebook AND the responses in the data BOTH in the original AND the processed data files

- Make sure you understand the instrument and the coding in the data files

<br>

**Ok, groups are you ready to start analyzing the polarization in your issue?**

<br>

Groups, your job today is to answer this research question:

- Does the data on your issue show an increase in partisan polarization over time?

<br>

SLIDE: Let's step through your analyses in a series of tasks

:::




## Polarization and the ANES {background-image="Images/background-white_house_graphic_blue_v2.png" .center}

<br>

1. Gay Adoption (VCF0878, p388)
    - "Yes"

2. Limit trade to protect US jobs (VCF9231, p600)
    - "Favor new limits"

3. Death penalty for murderers (VCF9237_v2, p603)
    - "Favor" (combines "not strongly" + "strongly")

4. Guns (VCF9238, p604)
    - "More difficult"

::: notes

STEP 1: Let's get a sense of the overall trend in your data

- e.g. is support/opposition for this issue changing across time?

<br>

GROUPS, visualize the proportion of responses selecting the option on the slide across time

- For group 1, what proportion of respondents are in favor of gay adoption in each year in the dataset?

- For group 4, what proportion of respondents are in favor of making gun purchases more difficult in each year in the dataset?

- **Make sense?**

<br>

SLIDE: To make sure you're on the right track let's check these results before moving to tougher tasks

:::



## {background-image="Images/background-white_house_graphic_blue_v2.png"}

::: {.panel-tabset}

### Gay Adoption

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
tab1a <- d |>
  filter(!is.na(VCF0878)) |>
  group_by(VCF0004) |>
  summarize(
    prop = mean(VCF0878 == "Yes")
  )

tab1a |>
  ggplot(aes(x = VCF0004, y = prop)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1992, 2020, 2)) +
  labs(x = "", y = "", title = "US Public Approval of Gay Adoption") +
  #ggrepel::geom_text_repel(aes(label = str_c(round(prop, 2)*100, "%")))
  geom_text(aes(label = str_c(round(prop, 2)*100, "%")), nudge_y = .05)
```

### Limit Trade

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
tab1b <- d |>
  filter(!is.na(VCF9231)) |>
  group_by(VCF0004) |>
  summarize(
    prop = mean(VCF9231 == "Favor new limits")
  )

tab1b |>
  ggplot(aes(x = VCF0004, y = prop)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1988, 2020, 2)) +
  labs(x = "", y = "", title = "US Public Approval of Limiting Trade") +
  #ggrepel::geom_text_repel(aes(label = str_c(round(prop, 2)*100, "%")))
  geom_text(aes(label = str_c(round(prop, 2)*100, "%")), nudge_y = .05)
```

### Death Penalty

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
tab1c <- d |>
  filter(!is.na(VCF9237_v2)) |>
  group_by(VCF0004) |>
  summarize(
    prop = mean(VCF9237_v2 == "Favor")
  )

tab1c |>
  ggplot(aes(x = VCF0004, y = prop)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1988, 2020, 2)) +
  labs(x = "", y = "", title = "US Public Approval of the Death Penalty") +
  #ggrepel::geom_text_repel(aes(label = str_c(round(prop, 2)*100, "%")))
  geom_text(aes(label = str_c(round(prop, 2)*100, "%")), nudge_y = .05)
```

### Guns

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
tab1d <- d |>
  filter(!is.na(VCF9238)) |>
  group_by(VCF0004) |>
  summarize(
    prop = mean(VCF9238 == "More difficult")
  )

tab1d |>
  ggplot(aes(x = VCF0004, y = prop)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1988, 2020, 2)) +
  labs(x = "", y = "", title = "US Public Approval for New Restrictions on Guns") +
  #ggrepel::geom_text_repel(aes(label = str_c(round(prop, 2)*100, "%")))
  geom_text(aes(label = str_c(round(prop, 2)*100, "%")), nudge_y = .05)
```

:::

::: notes

*Step through each and make sure they got these results PLUS understand what this shows*

<br>

Time to use this data to measure polarization!

- GROUPS, redo this exercise but create separate lines for the Republicans and the Democrats in the sample (VCF0303)

- **Questions?**

- Go!

:::



## {background-image="Images/background-white_house_graphic_blue_v2.png"}

::: {.panel-tabset}

### Gay Adoption

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
# Yes % by party ID
d |>
  filter(!is.na(VCF0878)) |>
  group_by(VCF0303, VCF0004) |>
  summarize(.groups = "keep",
    prop = mean(VCF0878 == "Yes")
  ) |>
  filter(!is.na(VCF0303)) |>
  ggplot(aes(x = VCF0004, y = prop, color = VCF0303)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1992, 2020, 2)) +
  labs(x = "", y = "", title = "Approval of Gay Adoption by Party Identification", color = "") +
  scale_color_manual(values = c("blue", "darkgrey", "red"))
```

### Limit Trade

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
d |>
  filter(!is.na(VCF9231)) |>
  group_by(VCF0303, VCF0004) |>
  summarize(.groups = "keep",
    prop = mean(VCF9231 == "Favor new limits")
  ) |>
  filter(!is.na(VCF0303)) |>
  ggplot(aes(x = VCF0004, y = prop, color = VCF0303)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1988, 2020, 2)) +
  labs(x = "", y = "", title = "US Public Approval of Limiting Trade by Party Identification", color = "") +
  scale_color_manual(values = c("blue", "darkgrey", "red"))

```

### Death Penalty

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
d |>
  filter(!is.na(VCF9237_v2)) |>
  group_by(VCF0303, VCF0004) |>
  summarize(.groups = "keep",
    prop = mean(VCF9237_v2 == "Favor")
  ) |>
  filter(!is.na(VCF0303)) |>
  ggplot(aes(x = VCF0004, y = prop, color = VCF0303)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1988, 2020, 2)) +
  labs(x = "", y = "", title = "US Public Approval of Death Penalty by Party Identification", color = "") +
  scale_color_manual(values = c("blue", "darkgrey", "red"))
```

### Guns

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
d |>
  filter(!is.na(VCF9238)) |>
  group_by(VCF0303, VCF0004) |>
  summarize(.groups = "keep",
    prop = mean(VCF9238 == "More difficult")
  ) |>
  filter(!is.na(VCF0303)) |>
  ggplot(aes(x = VCF0004, y = prop, color = VCF0303)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1988, 2020, 2)) +
  labs(x = "", y = "", title = "US Public Approval for New Restrictions on Guns by Party Identification", color = "") +
  scale_color_manual(values = c("blue", "darkgrey", "red"))
```

:::

::: notes

*Step through each and make sure they got these results*

<br>

GROUPS, now I want you to directly measure the polarization!

- Make me a single line that represents the difference between the proportion of Democrats and the proportion of Republicans 

- **Questions on the task?**

- Go!

:::



## {background-image="Images/background-white_house_graphic_blue_v2.png"}

::: {.panel-tabset}

### Gay Adoption

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
# Difference by party ID
tab2a <- d |>
  filter(!is.na(VCF0878)) |>
  group_by(VCF0303, VCF0004) |>
  summarize(.groups = "keep",
    prop = mean(VCF0878 == "Yes")
  ) |>
  filter(VCF0303 %in% c("Democrats", "Republicans")) |>
  pivot_wider(names_from = VCF0303, values_from = prop) |>
  mutate(
    difference = Democrats - Republicans,
    policy = "Gay Adoption"
  )

tab2a |>
  ggplot(aes(x = VCF0004, y = difference)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(0, .4), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1992, 2020, 2)) +
  labs(x = "", y = "Approval Differences",
       title = "Differences in Approval of Gay Adoption (Democrats - Republicans)") +
  geom_text(aes(label = str_c(round(difference, 2)*100, "%")), nudge_y = .02)
```

### Limit Trade

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
tab2b <- d |>
  filter(!is.na(VCF9231)) |>
  group_by(VCF0303, VCF0004) |>
  summarize(.groups = "keep",
    prop = mean(VCF9231 == "Favor new limits")
  ) |>
  filter(VCF0303 %in% c("Democrats", "Republicans")) |>
  pivot_wider(names_from = VCF0303, values_from = prop) |>
  mutate(
    difference = Democrats - Republicans,
    policy = "Limit Trade"
  )

tab2b |>
  ggplot(aes(x = VCF0004, y = difference)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(-.4, .2), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1988, 2020, 2)) +
  labs(x = "", y = "Approval Differences",
       title = "Differences in Approval of Limiting Trade (Democrats - Republicans)") +
  geom_text(aes(label = str_c(round(difference, 2)*100, "%")), nudge_y = .02)
```

### Death Penalty

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
tab2c <- d |>
  filter(!is.na(VCF9237_v2)) |>
  group_by(VCF0303, VCF0004) |>
  summarize(.groups = "keep",
    prop = mean(VCF9237_v2 == "Favor")
  ) |>
  filter(VCF0303 %in% c("Democrats", "Republicans")) |>
  pivot_wider(names_from = VCF0303, values_from = prop) |>
  mutate(
    difference = Democrats - Republicans,
    policy = "Death Penalty"
  )

tab2c |>
  ggplot(aes(x = VCF0004, y = difference)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(-.5, 0), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1988, 2020, 2)) +
  labs(x = "", y = "Approval Differences",
       title = "Differences in Approval of the Death Penalty (Democrats - Republicans)") +
  geom_text(aes(label = str_c(round(difference, 2)*100, "%")), nudge_y = .02)
```

### Guns

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
tab2d <- d |>
  filter(!is.na(VCF9238)) |>
  group_by(VCF0303, VCF0004) |>
  summarize(.groups = "keep",
    prop = mean(VCF9238 == "More difficult")
  ) |>
  filter(VCF0303 %in% c("Democrats", "Republicans")) |>
  pivot_wider(names_from = VCF0303, values_from = prop) |>
  mutate(
    difference = Democrats - Republicans,
    policy = "Gun Control"
  )

tab2d |>
  ggplot(aes(x = VCF0004, y = difference)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_y_continuous(limits = c(.1, .6), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1988, 2020, 2)) +
  labs(x = "", y = "Approval Differences",
       title = "Differences in Approval of New Restrictions on Guns (Democrats - Republicans)") +
  geom_text(aes(label = str_c(round(difference, 2)*100, "%")), nudge_y = .02)
```

:::

::: notes

*Step through each and make sure they got these results PLUS understand what this shows*

<br>

**BOTTOM LINE, how much partisan polarization are we seeing in these issues?**

- **And how much since 2007 (the end of the analyses in the Fiorina and Abrams)?**

<br>

SLIDE: What happened?

:::




## {background-image="Images/background-white_house_graphic_blue_v2.png"}

```{r, fig.retina=3, fig.align='center', fig.asp=.618, out.width="85%"}
rbind(tab2b, tab2c, tab2d) |>
  ggplot(aes(x = VCF0004, y = difference, color = policy)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  scale_y_continuous(limits = c(-.5, .5), labels = scales::percent_format()) +
  labs(x = "", y = "Differences in Policy Support (Dems - Repubs)",
       title = "Measuring Issue Polarization using ANES Data") +
  scale_x_continuous(breaks = seq(1988, 2020, 2)) +
  annotate("text", x = c(1990, 1990, 2001), y = c(-.25, .05, .23), label = c("Death Penalty", "Limit Trade", "Gun Control")) +
  guides(color = "none")
```

::: notes

**Any hypotheses as to what happened since the Fiorina and Abrams article to explain the growing issue polarization?**

:::



## For Next Class {background-image="Images/background-blue_triangles_flipped.png" .center}

<br>

Defining and Measuring Affective Polarization

1. Iyengar, Lelkes, Levendusky, Malhotra, & Westwood (2019)

2. Canvas: Find us an example of affective polarization in our current events. In other words, an expression of "animosity between the parties" that is NOT based on a difference in policy preferences.

::: notes

:::