---
format: revealjs
---

## Today's Agenda {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

```{r}
# background-size="1920px 1080px"
library(tidyverse)
library(readxl)

# Input ANES data
d <- read_excel("../../Data_to_Explore/ANES/ANES-1978-2020-Selected_Variables.xlsx", na = "NA", guess_max = 100000)
```

<br>

::: {.r-fit-text}

**II. How serious is the problem of affective polarization in the American electorate?**

- Getting hands-on with ANES data to examine polarization trends since 2008

:::

<br>

::: r-stack
Justin Leinaweaver (Fall 2024)
:::

::: notes
Prep for Class

1. Readings
    - (Homans and McFadden 2022 - NYT on how politics is ending friendships and harming relationships)
    - (Aldrich 2017 on ANES, e.g. what is it?)
    - (Review the ANES data extract and codebook: Make sure you understand what each variable represents and how they are coded)

:::



## For Today {background-image="Images/background-blue_triangles_flipped.png" .center}

<br>

Getting hands-on with ANES data to examine polarization trends since 2008

1. Homans and McFadden (2022)

2. Aldrich (2017) on ANES

3. Review the ANES data extract and codebook: Make sure you understand what each variable represents and how they are coded

::: notes


:::


## Is something changing? {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

Homans and McFadden 2022 - NYT on how politics is ending friendships and harming relationships



## What is the ANES? {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

Aldrich 2017 on ANES, e.g. what is it?



## The ANES Codebook {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

Review the ANES data extract and codebook: Make sure you understand what each variable represents and how they are coded

- Evaluate the survey designs
- What does this show us?



## The ANES Data: Prompt 1 {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

::: {.r-fit-text}
Count proportion party identification across time (omit don't knows)
:::

```{r, fig.retina=3, fig.asp=0.618, fig.align='center', fig.width=8.5}
# p116-117: Party Identification - VCF0303 (Party Identification - Summary 3 categories)
parties_time <- d |>
  group_by(Year) |>
  count(VCF0303) |>
  mutate(
    party = case_when(
      VCF0303 == 0 ~ "Don't know",
      VCF0303 == 1 ~ "Democrats",
      VCF0303 == 2 ~ "Independents",
      VCF0303 == 3 ~ "Republicans"
    )
  ) |>
  select(-VCF0303)

# # Table
# parties_time |>
#   pivot_wider(names_from = party, values_from = n)

## Version 1: Filled bar plot
parties_time |>
  filter(party != "Don't know") |>
  ggplot(aes(x = Year, y = n, fill = party)) +
  geom_col(position = "fill") +
  theme_bw() +
  labs(x = "", y = "", fill = "") +
  scale_fill_manual(values = c("dodgerblue", "bisque1", "firebrick2")) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1978, 2020, 2)) +
  guides(fill = "none") +
  annotate("text", x = 1978, y = .13, label = "Republican", srt = 90, color = "white") +
  annotate("text", x = 1978, y = .71, label = "Democratic", srt = 90, color = "white")
```



## The ANES Data: Prompt 1 {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

::: {.r-fit-text}
Count proportion party identification across time (omit don't knows)
:::

```{r, fig.retina=3, fig.asp=0.618, fig.align='center', fig.width=8}
# p116-117: Party Identification - VCF0303 (Party Identification - Summary 3 categories)

## Version 2: Line plot of percentages across time
parties_time |>
  filter(party != "Don't know") |>
  mutate(
    Total = sum(n),
    Proportion = n/Total
  ) |>
  ggplot(aes(x = Year, y = Proportion, color = party)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(x = "", y = "", fill = "") +
  scale_color_manual(values = c("dodgerblue", "tan1", "firebrick2")) +
  scale_y_continuous(limits = c(0, .65), breaks = seq(0, .6, .1), labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(1978, 2020, 2)) +
  guides(color = "none") +
  annotate("text", x = 1980, y = c(.6, .27, .08), label = c("Democratic", "Republican", "Independent"), color = c("dodgerblue", "firebrick2", "tan1"))
```



## The ANES Data: Prompt 2 {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

How much does each group like their own party (averages across time)? 

```{r, fig.retina=3, fig.asp=0.618, fig.align='center', fig.width=8}
# Split sample
reps <- subset(d, d$VCF0303 == 3)
dems <- subset(d, d$VCF0303 == 1)

# Calculate own feeling average
# VCF0224 - Republican Party
# VCF0218 - Democratic Party
x1 <- reps |>
  group_by(Year) |>
  summarize(
    Own_Avg = mean(VCF0224, na.rm = TRUE),
    Own_Med = median(VCF0224, na.rm = TRUE),
    Own_Weighted = weighted.mean(VCF0224, Weights, na.rm = TRUE)
  ) |>
  mutate(
    Group = "Republicans"
  )

x2 <- dems |>
  group_by(Year) |>
  summarize(
    Own_Avg = mean(VCF0218, na.rm = TRUE),
    Own_Med = median(VCF0218, na.rm = TRUE),
    Own_Weighted = weighted.mean(VCF0218, Weights, na.rm = TRUE)
  ) |>
  mutate(
    Group = "Democrats"
  )

rbind(x1, x2) |>
  ggplot(aes(x = Year, y = Own_Avg, color = Group)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(x = "", y = "Thermometer Scale (0-100)",
       title = "Party Identifiers' Self-Reflection") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = seq(1978, 2020, 2)) +
  scale_color_manual(values = c("dodgerblue","firebrick2")) +
  guides(color = "none")
```



## The ANES Data: Prompt 3 {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

How much does each group like the other party (averages across time)? 

```{r, fig.retina=3, fig.asp=0.618, fig.align='center', fig.width=8}
# Calculate other feeling average
# VCF0224 - Republican Party
# VCF0218 - Democratic Party
x1a <- reps |>
  group_by(Year) |>
  summarize(
    Other_Avg = mean(VCF0218, na.rm = TRUE),
    Other_Med = median(VCF0218, na.rm = TRUE),
    Other_Weighted = weighted.mean(VCF0218, Weights, na.rm = TRUE)
  ) |>
  mutate(
    Group = "Republicans"
  )

x2a <- dems |>
  group_by(Year) |>
  summarize(
    Other_Avg = mean(VCF0224, na.rm = TRUE),
    Other_Med = median(VCF0224, na.rm = TRUE),
    Other_Weighted = weighted.mean(VCF0224, Weights, na.rm = TRUE)
  ) |>
  mutate(
    Group = "Democrats"
  )

rbind(x1a, x2a) |>
  ggplot(aes(x = Year, y = Other_Avg, color = Group)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(x = "", y = "Thermometer Scale (0-100)",
       title = "Party Identifiers' Feelings About the Other Party") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_continuous(breaks = seq(1978, 2020, 2)) +
  scale_color_manual(values = c("dodgerblue","firebrick2")) +
  guides(color = "none")
```

::: notes



:::



## The ANES Data: Prompt 4 {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

Is the difference growing or shrinking across time? e.g. own party average minus rival party average feelings across time

```{r, fig.retina=3, fig.asp=0.618, fig.align='center', fig.width=8}
# Calculate other feeling average
# VCF0224 - Republican Party
# VCF0218 - Democratic Party
x4a <- reps |>
  group_by(Year) |>
  summarize(
    Self_Avg = mean(VCF0224, na.rm = TRUE),
    Other_Avg = mean(VCF0218, na.rm = TRUE),
    Difference = Self_Avg - Other_Avg
  ) |>
  mutate(
    Group = "Republicans"
  )

x4b <- dems |>
  group_by(Year) |>
  summarize(
    Self_Avg = mean(VCF0218, na.rm = TRUE),
    Other_Avg = mean(VCF0224, na.rm = TRUE),
    Difference = Self_Avg - Other_Avg
  ) |>
  mutate(
    Group = "Democrats"
  )

rbind(x4a, x4b) |>
  ggplot(aes(x = Year, y = Difference, color = Group)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(x = "", y = "Thermometer Scale (0-100)",
       title = "The Difference Between Approval of the Self and Other") +
  scale_y_continuous(limits = c(20, 55)) +
  scale_x_continuous(breaks = seq(1978, 2020, 2)) +
  scale_color_manual(values = c("dodgerblue","firebrick2")) +
  guides(color = "none")
```

::: notes

Next slide: underlying averages

:::


## The ANES Data: Prompt 4 {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

Is the difference growing or shrinking across time? e.g. own party average minus rival party average feelings across time

```{r, fig.retina=3, fig.asp=0.618, fig.align='center', fig.width=8}
# Calculate other feeling average
# VCF0224 - Republican Party
# VCF0218 - Democratic Party
combo4 <- rbind(x4a, x4b)

combo4_pre <- subset(combo4, Year <= 2002)
combo4_post <- subset(combo4, Year >= 2002)

combo4 |>
  ggplot(aes(x = Year, y = Difference, color = Group)) +
  geom_point() +
  geom_smooth(data = combo4_pre, method = "lm", se = FALSE) +
  geom_smooth(data = combo4_post, method = "lm", se = FALSE) +
  theme_bw() +
  labs(x = "", y = "Thermometer Scale (0-100)",
       title = "Something happened in 2000...") +
  scale_y_continuous(limits = c(20, 55)) +
  scale_x_continuous(breaks = seq(1978, 2020, 2)) +
  scale_color_manual(values = c("dodgerblue","firebrick2")) +
  guides(color = "none")
```

::: notes


:::



## The ANES Data: Prompt 5 {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

Are the effects more dramatic for the politically active? e.g. "Voter" variable I created that asks if the respondent voted in the last election

```{r, fig.retina=3, fig.asp=0.618, fig.align='center', fig.width=8}
x5a <- reps |>
  filter(!is.na(Voter)) |>
  group_by(Voter, Year) |>
  summarize(
    Self_Avg = mean(VCF0224, na.rm = TRUE),
    Other_Avg = mean(VCF0218, na.rm = TRUE),
    Difference = Self_Avg - Other_Avg
  ) |>
  mutate(
    Group = "Republicans"
  )

x5b <- dems |>
  filter(!is.na(Voter)) |>
  group_by(Voter, Year) |>
  summarize(
    Self_Avg = mean(VCF0218, na.rm = TRUE),
    Other_Avg = mean(VCF0224, na.rm = TRUE),
    Difference = Self_Avg - Other_Avg
  ) |>
  mutate(
    Group = "Democrats"
  )

rbind(x5a, x5b) |>
  ggplot(aes(x = Year, y = Difference, color = Voter)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(x = "", y = "Differences (Thermometer Scale, 0-100)") +
  scale_y_continuous(limits = c(10, 60), breaks = seq(10, 60, 10)) +
  scale_x_continuous(breaks = seq(1978, 2020, 4)) +
  scale_color_manual(values = c("darkorange1","forestgreen")) +
  theme(legend.position = "bottom") +
  facet_wrap(~ Group)
```

::: notes


:::



## The ANES Data: Prompt 6 {background-image="Images/background-white_house_graphic_v4_muted_text2.png" .center}

Are the effects more dramatic for various confounders? e.g. age, gender, race/ethnicity, education, region, knowledge

```{r, fig.retina=3, fig.asp=0.618, fig.align='center', fig.width=8}
# Various confounders to consider (taken from the ANES visualization tool): Age - Recoded from VCF0103 (Respondent - Age); Gender - VCF0104 (Respondent - Gender); Race/Ethnicity - VCF0105b (Race-Ethnicity Summary, 4 categories); Education - Recoded from VCF0110 (Respondent - Education 4 categories); Region - VCF0113 (Political South/Nonsouth); Party Identification - VCF0303 (Party Identification - Summary 3 categories); Knowledge - VCF0729 (Party with House majority before Election?)

# Gender
x5a <- reps |>
  filter(!is.na(VCF0104), VCF0104 != 3, VCF0104 != 0) |>
  group_by(VCF0104, Year) |>
  summarize(
    Self_Avg = mean(VCF0224, na.rm = TRUE),
    Other_Avg = mean(VCF0218, na.rm = TRUE),
    Difference = Self_Avg - Other_Avg
  ) |>
  mutate(
    Group = "Republicans"
  )

x5b <- dems |>
  filter(!is.na(VCF0104), VCF0104 != 3, VCF0104 != 0) |>
  group_by(VCF0104, Year) |>
  summarize(
    Self_Avg = mean(VCF0218, na.rm = TRUE),
    Other_Avg = mean(VCF0224, na.rm = TRUE),
    Difference = Self_Avg - Other_Avg
  ) |>
  mutate(
    Group = "Democrats"
  )

rbind(x5a, x5b) |>
  ggplot(aes(x = Year, y = Difference, color = factor(VCF0104))) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(x = "", y = "Differences (Thermometer Scale, 0-100)") +
  scale_y_continuous(limits = c(10, 60), breaks = seq(10, 60, 10)) +
  scale_x_continuous(breaks = seq(1978, 2020, 4)) +
  scale_color_manual(values = c("darkorange1","forestgreen")) +
  theme(legend.position = "bottom") +
  facet_wrap(~ Group)

```

::: notes

- Check your results while adjusting for the confounder

- Omitted: Repeat prompts 1-4 using ideology not party

:::






## For Next Class {background-image="Images/background-blue_triangles_flipped.png" .center}

<br>

Defining and Measuring Affective Polarization

1. Iyengar, Lelkes, Levendusky, Malhotra, & Westwood (2019)

2. ASSIGNMENT - Find us an example of affective polarization in action with the current election. In other words, an expression of "animosity between the parties" that is NOT based on a difference in policy preferences.


::: notes

:::